\documentclass[11pt]{article}
\usepackage[margin=2.4cm]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks=true,linkcolor=blue!50!black,urlcolor=blue!50!black}

\newcommand{\us}{u_s}
\newcommand{\un}{u_n}
\newcommand{\Us}{\bar U_s}
\newcommand{\ds}{\partial_s}
\newcommand{\dn}{\partial_n}
\newcommand{\dt}{\partial_t}
\newcommand{\code}[1]{\texttt{\small #1}}

\title{\bf Derivation of every equation in \code{sw\_sn\_driver.py}\\
\large Full shallow water on a finite-amplitude meander, channel-fitted $(s,n)$}
\author{\code{numerical/dedalus\_meander\_full\_SW}}
\date{}

\begin{document}
\maketitle
\vspace{-8mm}

\noindent This note derives, in order, every formula that appears in
\code{sw\_sn\_driver.py}. Each section names the code symbol it produces.

\tableofcontents

% ===================================================================
\section{Why $(s,n)$, and the metric $\sigma=1+nC$}
\label{sec:metric}

The channel meanders with \emph{finite} amplitude, so its banks are not straight
lines and a Chebyshev grid cannot be laid on them directly. We therefore use
\textbf{channel-fitted} coordinates: $s$ = arc length along the centreline,
$n$ = signed distance across it. In $(s,n)$ the channel is a straight strip
$n\in[-b,b]$ whatever the planform does.

Let the centreline be $\bm r_c(s)$ with unit tangent $\hat{\bm t}(s)$ and unit
normal $\hat{\bm n}(s)$, and let $C(s)$ be its curvature. A point of the domain is
\begin{equation}
  \bm r(s,n)=\bm r_c(s)+n\,\hat{\bm n}(s).
\end{equation}
With the Frenet relation $\mathrm d\hat{\bm n}/\mathrm ds=+C\,\hat{\bm t}$
(this \emph{defines} our sign convention: $n$ increases \emph{away} from the centre
of curvature; equivalently the direction angle is $\theta(s)=-\int^s C\,\mathrm ds'$),
\begin{equation}
  \frac{\partial\bm r}{\partial s}=\hat{\bm t}+n\frac{\mathrm d\hat{\bm n}}{\mathrm ds}
  =(1+nC)\,\hat{\bm t},
  \qquad
  \frac{\partial\bm r}{\partial n}=\hat{\bm n},
\end{equation}
so the coordinates are orthogonal with scale factors
\begin{equation}
  \boxed{\;h_s=\sigma\equiv 1+nC(s),\qquad h_n=1.\;}
  \tag{$\to$ \code{sigma\_metric}, \code{sig}}
\end{equation}
$\sigma$ is the local stretching of an along-channel line at offset $n$: the outer
bank ($n$ with the sign of $C$) is longer, the inner bank shorter.

\paragraph{Validity.} $\sigma>0$ is required or the mapping folds. Since
$|n|\le b$, this is $|C|b<1$: the bend radius must exceed the half-width. The test
\code{tests/test\_base\_profiles.py} asserts $\min\sigma>0$.

\paragraph{Sign convention (important).} Mirroring the channel ($C\to-C$) maps the
problem to itself with $n\to-n$, so the linear growth rate $\sigma(k)$ and phase
speed $c(k)$ are \emph{unchanged}; only the labelling of ``outer'' vs ``inner''
bank flips. The convention above is the one used by the solver
($\sigma=1+nC$); the lab-frame renderer must integrate $\theta=-\int C\,\mathrm ds$
to match it.

% ===================================================================
\section{Shallow-water equations in $(s,n)$}

\subsection{The starting (Cartesian) system}
Non-rotating depth-averaged shallow water over a bed of still-water depth $H$,
free surface $\eta$, total depth $h=\eta+H$:
\begin{equation}
  \dt\bm u+(\bm u\cdot\nabla)\bm u=-g\nabla\eta+\bm F,
  \qquad
  \dt\eta+\nabla\cdot(h\bm u)=0 .
  \label{eq:sw-cart}
\end{equation}
The pressure is hydrostatic, $p=\rho g(\eta-z)$, so the horizontal pressure
gradient is exactly $-g\nabla\eta$; this is the $L\gg H$ (long-wave) limit.
Keeping $\dt\eta$ is what admits \textbf{gravity waves}; the rigid-lid packages in
\code{../} drop it and are left with the vortical branch only.

\subsection{Curvilinear operators}
For orthogonal coordinates with scale factors $(h_s,h_n)=(\sigma,1)$:
\begin{align}
  \nabla\phi&=\frac{1}{\sigma}\ds\phi\,\hat{\bm t}+\dn\phi\,\hat{\bm n},\\
  \nabla\cdot\bm F&=\frac{1}{\sigma}\Big[\ds F_s+\dn(\sigma F_n)\Big],
  \label{eq:div}\\
  \zeta=(\nabla\times\bm u)\cdot\hat{\bm z}&=\frac{1}{\sigma}\Big[\ds \un-\dn(\sigma \us)\Big].
  \label{eq:vort}
\end{align}
The material acceleration acquires \emph{metric} (centrifugal/Coriolis-like) terms
from the rotation of the basis vectors. With
$\mathrm D/\mathrm Dt=\dt+(\us/\sigma)\ds+\un\dn$ and
$\partial h_s/\partial n=C$, $\partial h_n/\partial s=0$, the standard result is
\begin{equation}
  a_s=\frac{\mathrm D\us}{\mathrm Dt}+\frac{C\,\us\un}{\sigma},
  \qquad
  a_n=\frac{\mathrm D\un}{\mathrm Dt}-\frac{C\,\us^{2}}{\sigma}.
  \label{eq:accel}
\end{equation}

\subsection{The system solved}
Substituting \eqref{eq:div}--\eqref{eq:accel} into \eqref{eq:sw-cart}:
\begin{align}
  &\dt\eta+\frac{1}{\sigma}\Big[\ds(h\us)+\dn(\sigma h\un)\Big]=0,
  \label{eq:cont}\\
  &\dt \us+\frac{\us}{\sigma}\ds \us+\un\dn \us
   +\underbrace{\frac{C\us\un}{\sigma}}_{\text{bend coupling}}
   =-\frac{g}{\sigma}\ds\eta-r_s\us+\nu\nabla^2\us,
  \label{eq:smom}\\
  &\dt \un+\frac{\us}{\sigma}\ds \un+\un\dn \un
   \underbrace{-\frac{C\us^{2}}{\sigma}}_{\text{\bf centrifugal}}
   =-g\,\dn\eta-r_n\un+\nu\nabla^2\un.
  \label{eq:nmom}
\end{align}
The term $-C\us^2/\sigma$ is the whole reason for going to $(s,n)$: it is the
centrifugal acceleration of water rounding a bend, and it is what tilts the free
surface up against the outer bank (\S\ref{sec:base}). No straight-channel model
contains it.

% ===================================================================
\section{Non-dimensionalisation: where the Froude number enters}
\label{sec:nondim}

Scale lengths by the half-width $b$, speeds by the centreline speed
$U_c=U_0+\Delta$, time by $b/U_c$, and depths/$\eta$ by the mean depth $H_0$.
The only place $g$ survives is the pressure gradient, where it appears as
\begin{equation}
  \frac{gH_0}{U_c^{2}}=\frac{1}{F^{2}},\qquad
  \boxed{\;F\equiv\frac{U_c}{\sqrt{gH_0}},\qquad g_{\rm eff}=\frac1{F^{2}}\;}
  \tag{$\to$ \code{g\_eff}}
\end{equation}
so the non-dimensional gravity-wave speed is $\sqrt{g_{\rm eff}H_0}=1/F$.

\paragraph{Why $F$ must be specified --- and why sweeping it settles nothing.}
$F$ is an \emph{input} (a property of the river), exactly like a Reynolds number:
without it the system \eqref{eq:cont}--\eqref{eq:nmom} is not defined. At any $F$
\emph{both} families exist --- gravity waves at speed $1/F$ and the vortical
branch at $O(U)$. It is tempting to identify the meander by varying $F$ and noting
that $\sigma$ and $c$ do not move. That inference is \textbf{invalid}, and we do not
make it: as \S\ref{sec:notcontain}(iii) shows, writing $P\equiv g\eta$ leaves the
momentum equations $F$-free, so $F$ survives only as $F^{2}$ on the continuity
time-derivative --- a pure divergence parameter. \emph{Any} slow balanced mode is
$F$-insensitive by scaling, so the observed insensitivity confirms the rigid-lid
reduction rather than identifying a wave family. The question is settled instead by
the energetics of \S\ref{sec:classify}, not by a Froude sweep.

% ===================================================================
\section{The base state}
\label{sec:base}

\subsection{Parabolic jet and the constant channel-$\beta$}
\begin{equation}
  \boxed{\;\Us(n)=U_0+\Delta\Big(1-\frac{n^{2}}{b^{2}}\Big)\;}
  \tag{$\to$ \code{ubar\_s}}
\end{equation}
so $\Us(0)=U_0+\Delta$ (centre) and $\Us(\pm b)=U_0$ (bank edge), and
\begin{equation}
  \dn\Us=-\frac{2\Delta n}{b^{2}}\ \ (\to\code{ubar\_s\_n}),\qquad
  \dn^{2}\Us=-\frac{2\Delta}{b^{2}}=\text{const}\ \ (\to\code{ubar\_s\_nn}).
\end{equation}
This constant is the load-bearing property. From \eqref{eq:vort} with $\un=0$ and
(for the moment) $C=0$, the base vorticity is $\bar\zeta=-\dn\Us$, hence
\begin{equation}
  \boxed{\;\frac{\partial\bar\zeta}{\partial n}=-\dn^{2}\Us=\frac{2\Delta}{b^{2}}
  =\text{constant}\;}
  \label{eq:beta}
\end{equation}
--- exactly constant for a \emph{straight} channel. At finite curvature the exact
gradient carries two further terms, $\bar q_n=-\dn^2\Us+\kappa\,\dn\Us/\sigma
+\kappa^{2}\Us/\sigma^{2}$ with $\kappa=-\bar C$ (verified against an independently
derived channel-following QGPV note in \code{tests/test\_vs\_qgpv\_doc.py}). At the bank
sinuosity used by the runs that carry the conclusions these add $1.8\%$; at the most
sinuous run in the study, $16.6\%$. One consequence matters for \S\ref{sec:results}:
setting $\Delta=0$ leaves the $\kappa^{2}\Us/\sigma^{2}$ term, so it removes the
gradient only to $O(\kappa^{2}\Us)$ --- a residual of $0.011\%$ of the sheared gradient
at the sinuosity the controls use, and therefore negligible there, but not literally zero.
--- a uniform cross-channel gradient of background vorticity, which is why the jet is
taken parabolic rather than, say, sinusoidal. It is tempting to call this a channel
$\beta$ and conclude that a \emph{vortical (Rossby-type) wave} is thereby possible;
that inference does \textbf{not} follow, and we do not make it. A planetary $\beta$
supports discrete Rossby modes because the domain is unbounded or the PV profile has
an edge; a smooth, single-signed $\dn\bar q$ between two walls has no PV edge, and by
the classical result of Case (1960) it supports only a \emph{continuous} spectrum ---
sheared disturbances that decay algebraically, not an exponentially growing normal
mode. Discrete modes require a PV discontinuity or extremum (Bretherton 1966;
Rayleigh's inflexion criterion), which \eqref{eq:beta} explicitly excludes. What
$\Delta$ does give us is a clean control knob: $\Delta$ is the shear, so setting
$\Delta=0$ removes the ``$\beta$'' entirely and tests directly whether the growing
meander needs it (\S\ref{sec:notcontain}). (For $C\neq0$ the same calculation gives
$\bar\zeta=-\dn\Us-C\Us/\sigma$, a curvature correction.)

\subsection{Base curvature = the finite background meander}
\begin{equation}
  \boxed{\;\bar C(s)=\bar C_{\rm amp}\cos(k_m s),\qquad
  \bar C_{\rm amp}=A_{\rm bank}k_m^{2}\ \text{(default)}\;}
  \tag{$\to$ \code{cbar}}
\end{equation}
For a centreline of lateral amplitude $A_{\rm bank}$ and wavenumber $k_m$, the
small-slope curvature is $-\partial_s^2(A_{\rm bank}\cos k_m s)=A_{\rm bank}k_m^{2}\cos k_m s$;
\code{Cbar\_amp} overrides this to set a genuinely tight bend directly.
\textbf{$A_{\rm bank}$ is the fixed background channel shape --- it is \emph{not} the
perturbation seed $A_0$ of \S\ref{sec:seed}, and it does not evolve.}

\subsection{Superelevation}
For a steady base with $\bar U_n=0$ and $\ds(\cdot)=0$, the $n$-momentum
\eqref{eq:nmom} reduces to a balance between the centrifugal term and the pressure
gradient:
\begin{equation}
  \boxed{\;g\,\dn\bar\eta=\frac{\bar C\,\Us^{2}}{\bar\sigma}\;}
  \tag{$\to$ \code{etabar}}
  \label{eq:super}
\end{equation}
integrated as $\bar\eta(s,n)=\dfrac{\bar C(s)}{g}\displaystyle\int_0^n
\frac{\Us^{2}(n')}{1+n'\bar C(s)}\,\mathrm dn'$ (zero on the centreline). The water
surface tilts \emph{up towards the outer bank} --- the superelevation. Note
$\bar C\to0\Rightarrow\bar\eta\to0$ (asserted in the test).

\subsection{Discharge conservation}
The base continuity \eqref{eq:cont} with $\bar U_n=0$ requires
$\ds(\bar h\Us)=0$: the transport is the same through every cross-section. With a
cross-channel-only bed this is automatic ($\Us=\Us(n)$).

\paragraph{What sustains the base.} A parallel jet with bottom drag is not a
steady solution unless something drives it (the downhill slope). As in every
frozen-jet stability analysis we \emph{prescribe} $\Us$ and absorb the residual into
an implicit steady forcing; it cancels from the perturbation equations below.

% ===================================================================
\section{Linearisation}
\label{sec:lin}

Write $\us=\Us+\us'$, $\un=\un'$, $\eta=\bar\eta+\eta'$, $h=\bar h+\eta'$,
$C=\bar C+C'$, and drop products of primes.

\subsection{Continuity}
$\ds(h\us)=\ds(\bar h\Us)+\bar h\ds\us'+\Us\ds\eta'$, where the first term vanishes
(discharge), and $\dn(\sigma h\un)\to\dn(\bar\sigma\bar h\,\un')$. Hence
\begin{equation}
  \bar\sigma\,\dt\eta'+\bar h\,\ds\us'+\Us\,\ds\eta'+\dn\!\big(\bar\sigma\bar h\,\un'\big)=0,
  \tag{$\to$ eq.\ 1 in \code{build\_ivp\_SW}}
\end{equation}
already multiplied by $\bar\sigma$ (\S\ref{sec:sigma}).

\subsection{$s$-momentum}
$\frac{\us}{\sigma}\ds\us\to\frac{\Us}{\bar\sigma}\ds\us'$;
$\un\dn\us\to\un'\dn\Us$;
$\frac{C\us\un}{\sigma}\to\frac{\bar C\Us\un'}{\bar\sigma}$. Multiplying by
$\bar\sigma$ collects the two $\un'$ terms into one coefficient:
\begin{equation}
  \bar\sigma\dt\us'+\Us\ds\us'
  +\underbrace{\big(\bar\sigma\,\dn\Us+\bar C\,\Us\big)}_{\to\ \code{coefUn}}\un'
  +g\,\ds\eta'+\bar\sigma r_s\us'-\bar\sigma\nu\nabla^{2}\us'=0 .
  \tag{$\to$ eq.\ 2}
\end{equation}
The $\un'$ coefficient is the cross-channel advection of background momentum ---
proportional to the base vorticity --- and is the term that carries the
Rossby-type restoring.

\subsection{$n$-momentum and the curvature feedback}
\label{sec:feedback}
Expanding $-C\us^{2}/\sigma$ with $C=\bar C+C'$, $\sigma=\bar\sigma+nC'$:
\begin{equation}
  -\frac{C\us^{2}}{\sigma}\;\simeq\;
  \underbrace{-\frac{\bar C\Us^{2}}{\bar\sigma}}_{\text{base}}
  \;\underbrace{-\frac{2\bar C\Us\us'}{\bar\sigma}}_{\to\ \code{twoCbUb}}
  \;\underbrace{-\frac{C'\Us^{2}}{\bar\sigma}}_{\text{\bf new bend}}
  \;+\;\frac{nC'\,\bar C\Us^{2}}{\bar\sigma^{2}} .
\end{equation}
The perturbed pressure term contributes $-\bar\sigma g\dn\eta'-nC'g\dn\bar\eta$, and
using the base balance \eqref{eq:super}, $g\dn\bar\eta=\bar C\Us^2/\bar\sigma$, the
last piece is $-nC'\bar C\Us^{2}/\bar\sigma$. Adding it to the $+nC'\bar C\Us^2/\bar\sigma^2$
above and using $\bar\sigma-n\bar C=1$ leaves the remarkably clean forcing
\begin{equation}
  \boxed{\;\text{$n$-momentum forcing}=\frac{C'\,\Us^{2}}{\bar\sigma}\;\simeq\;C'\Us^{2}.\;}
\end{equation}
So the final equation, with the forcing moved to the left and
$C'=-\partial_s^{2}\zeta_c$ (\S\ref{sec:erosion}):
\begin{equation}
  \bar\sigma\dt\un'+\Us\ds\un'-2\bar C\Us\us'+\bar\sigma g\dn\eta'
  +\bar\sigma r_n\un'-\bar\sigma\nu\nabla^{2}\un'
  +\underbrace{\Us^{2}\,\partial_s^{2}\zeta_c}_{\text{bend $\to$ flow}}=0 .
  \tag{$\to$ eq.\ 3}
\end{equation}
\emph{This is the coupling that closes the meander instability}: a bend creates a
centrifugal forcing, which drives the flow, which erodes the bank, which deepens
the bend.

% ===================================================================
\section{Why every equation is multiplied by $\sigma$}
\label{sec:sigma}

The raw equations carry $1/\sigma=1/(1+nC)$, a \emph{rational} function of $n$.
Dedalus builds such a coefficient as a dense operator (its documentation warns
about exactly this), destroying the banded structure. Multiplying each equation
through by $\sigma$ removes every $1/\sigma$ and leaves coefficients that are
polynomial in $n$ (products of $\sigma$, $\bar h$, $\Us$), so the matrices stay
banded. Multiplying an equation by a nonzero function is an exact,
solution-preserving row operation.
The one residual $1/\bar\sigma$ (in the $C'$ forcing) is dropped at
$O(n\bar C\,C')$, i.e.\ second order in the perturbation amplitude \emph{and} the
base curvature.

% ===================================================================
\section{Bottom friction: the factor of two, and the free-surface term}

Quadratic (Ch\'ezy) drag is $\bm F=-\dfrac{C_f}{h}|\bm u|\bm u$. Two things must be
linearised: the \emph{velocity} in $|\bm u|\bm u$, and the \emph{depth} in $1/h$.
With $\bm u=(\Us+\us',\,\un')$, $h=\bar h+\eta'$ and $|\bm u|\simeq\Us+\us'$:
\begin{equation}
  -\frac{C_f|\bm u|\us}{h}
  \simeq\underbrace{-\frac{C_f\Us^{2}}{\bar h}}_{\text{base}}
  \underbrace{-\frac{2C_f\Us}{\bar h}\us'}_{r_s\ (\to\code{rs\_a})}
  \;\underbrace{+\;\frac{C_f\Us^{2}}{\bar h^{2}}\,\eta'}_{\text{\bf superelevation drag}\ (\to\code{reta\_a})},
  \qquad
  -\frac{C_f|\bm u|\un}{h}\simeq-\underbrace{\frac{C_f\Us}{\bar h}}_{r_n}\un' .
\end{equation}
The drag is \textbf{anisotropic}: streamwise perturbations are damped twice as fast
as cross-stream ones, because they perturb $|\bm u|$ as well as the direction.

The last term is Ikeda--Parker--Sawai's \emph{superelevation-drag} term
\cite{ikeda1981}: where the surface is raised the column is deeper, the drag
$\propto1/h$ is weaker, and the flow there accelerates. \textbf{It is the only route
by which the free surface drives bend flow, and hence the only place the Froude
number enters classical bend theory.} Omitting it (as an earlier version of this
code did) removes the physical mechanism this package exists to add, and makes any
Froude sensitivity test vacuous. The base part $C_f\Us^2/\bar h$ belongs to the
implicit driving balance and cancels.

% ===================================================================
\section{Boundary conditions, and why there are four tau terms}

The banks sit at fixed $n=\pm b$ in channel coordinates. Water cannot cross a
solid bank, so
\begin{equation}
  \boxed{\;\un'(\pm b)=0\;}\qquad\text{(no penetration)} .
\end{equation}
This assumes \emph{nothing} about the interior divergence: $\nabla\cdot\bm u\neq0$
everywhere inside (that is what the free surface is for). In particular the bank is
\emph{not} treated as a material streamline --- that would be a rigid-lid
(non-divergent) assumption and is inconsistent with \eqref{eq:cont}.

Lateral eddy viscosity $\nu\nabla^{2}$ makes each momentum equation second order in
$n$, which (i) is physically reasonable for a river, (ii) makes the parabolic jet an
actual steady solution, and (iii) gives a standard tau closure. A second-order
operator needs one boundary condition per wall per component, so with free slip
\begin{equation}
  \dn\us'(\pm b)=0\qquad\text{(free slip)} ,
\end{equation}
the tau budget is $2$ per viscous velocity component:
\[
  \underbrace{2}_{\us'}+\underbrace{2}_{\un'}+\underbrace{0}_{\eta'}=4
  \qquad(\to\ \code{t\_us1,t\_us2,t\_un1,t\_un2}),
\]
$\eta'$ needing none because its equation contains no second $n$-derivative.
Counting unknowns and equations:
$\{\us',\un',\eta',\zeta_c\}+4\ \text{tau}=8$ and
$3\ \text{PDE}+2\ \text{no-penetration}+2\ \text{free-slip}+1\ \text{erosion}=8$.
The tau terms are lifted onto the last two Chebyshev modes,
\code{lift(t,-1)+lift(t,-2)} on \code{ybasis.derivative\_basis(2)}.

% ===================================================================
\section{Erodible bank: the meander equation}
\label{sec:erosion}

The bank is a sediment boundary, not a material line: it recedes because the flow
erodes it. Ikeda's closure makes the migration rate proportional to the
\emph{near-bank excess longitudinal velocity}. For a \textbf{bend} the outer bank is
fast and the inner bank slow, so it is the \textbf{antisymmetric} part that moves the
centreline:
\begin{equation}
  \boxed{\;\dt\zeta_c=E\cdot\tfrac12\big[\us'(+b)-\us'(-b)\big],\qquad E=\varepsilon\,U_0\;}
  \label{eq:erosion}
\end{equation}
where $\zeta_c(s,t)$ is the lateral offset of the centreline (constant width).

\paragraph{Why antisymmetric --- and a correction.} The antisymmetric combination is
not an extra modelling choice: it follows from taking each bank's \emph{outward}
normal. The $+b$ bank's outward normal is $+\hat{\bm n}$ and the $-b$ bank's is
$-\hat{\bm n}$, so outward erosion of each bank is
$\dt\xi_+=+E\us'(+b)$ and $\dt\xi_-=-E\us'(-b)$, whence the centreline
$\tfrac12(\xi_++\xi_-)$ obeys exactly \eqref{eq:erosion}, and the width mode
$\tfrac12(\xi_+-\xi_-)$ obeys $\dt(\cdot)=\tfrac{E}{2}[\us'(+b)+\us'(-b)]$, which
vanishes identically because the response $\us'$ is \emph{odd} in $n$.
\emph{Correction to an earlier version of this note:} it claimed that
``independent banks'' would instead drive the centreline through the
\emph{symmetric} part and give $\sigma<0$. That was wrong --- it used $+\hat{\bm n}$
for both banks. The two formulations are \textbf{identical}, not opposed; the
earlier $\sigma<0$ came from that sign error killing the feedback, not from parity.

\paragraph{Closing the loop.} The migrating centreline changes the curvature,
\begin{equation}
  C'(s,t)=-\partial_s^{2}\zeta_c ,
\end{equation}
which re-enters the $n$-momentum as the forcing derived in \S\ref{sec:feedback}.
The cycle
\[
  \zeta_c\ \longrightarrow\ C'\ \longrightarrow\ \text{flow}\ \longrightarrow\
  \us'(\pm b)\ \longrightarrow\ \dt\zeta_c
\]
is the meander instability. Its growth rate is what the IVP measures.

% ===================================================================
\section{Seeding, and why the seed must be a domain mode}
\label{sec:seed}

The base state is steady \emph{by construction of the forcing}: started there, nothing ever happens. (It is not an unforced exact solution --- when $\bar C$ varies along $s$ the base $s$-momentum retains $-(g/\sigma)\partial_s\bar\eta\neq0$, the shallow-water analogue of the $\kappa_sU^2/h^3$ imbalance derived in the QGPV note. As in \S\ref{sec:lin}, that residual is taken to be balanced by an implicit mean forcing which cancels from the perturbation equations.)
Stability analysis therefore \emph{requires} a perturbation, and its growth rate is
the answer. Because the dynamics are linear, \emph{any} perturbation is admissible.

\paragraph{Broadband seeding (the default).} We seed the centreline with all
resolvable modes at once, equal amplitude and fixed pseudo-random phases:
\begin{equation}
  \zeta_c(s,0)=A_0\sum_{m=1}^{N_s/2-1}\cos\!\big(k_m s+\phi_m\big),
  \qquad k_m=\frac{2\pi m}{L_s},\qquad \us'=\un'=\eta'=0 .
\end{equation}
For a \emph{straight} base channel the coefficients are $s$-independent, so the
$s$-Fourier modes decouple exactly: each $k_m$ then evolves on its own and a
\textbf{single run contains the entire dispersion relation}, recovered by
demodulating each mode ($\to$ \code{per\_mode\_dispersion}). This is both cheaper
and safer than one run per wavelength --- in particular it removes any question of
whether the seeded wavenumber is commensurate with the domain. Runs are accordingly
identified by their physics --- bed $H$, initial bank sinuosity $\bar C_{\rm amp}b$,
friction $C_f$, and the base speeds $U_0,\Delta$ ($\to$ \code{run\_tag}) --- never by
a seeded wavelength. $\Delta$ belongs in that label even though it reads as a mere
``speed'': by \eqref{eq:beta} it \emph{is} the cross-channel vorticity gradient, so
$\Delta=0$ (plug) and $\Delta<0$ (reversed shear) are the two sharpest experiments in
the study, and omitting it would let them collide.
At finite bank sinuosity the coefficients are $s$-periodic, the
modes couple (Floquet), and the per-mode numbers are Bloch quantities rather than
eigenvalues.

\paragraph{Convergence gate.} A fitted growth rate is meaningless unless the mode
actually grew: we record the e-foldings achieved over the fit window and mark any
mode with fewer than three as not converged. (Without this gate, slowly-evolving
modes return fitted rates whose \emph{sign} depends on the fit window.)
\paragraph{$A_0$ is a free constant.} The system is linear, so if
$(\us',\un',\eta',\zeta_c)$ solves it then so does $\lambda\times$ it. Only
$\sigma$, $c$ and the \emph{ratios} between fields are physical; the overall size is
set by whatever initial disturbance one assumes. This is why the movie may fix the
overall constant once (final meander $=0.5b$) and apply it to every frame and every
field --- that is a re-choice of $A_0$, not a distortion.

% ===================================================================
\section{Measuring $\sigma$ and $c$}

The centreline is demodulated onto its seeded mode,
$a(t)=\widehat{\zeta_c}(m,t)$, and
\begin{equation}
  \sigma=\frac{\mathrm d}{\mathrm dt}\ln|a|,\qquad
  c=-\frac1k\frac{\mathrm d}{\mathrm dt}\arg a ,
  \tag{$\to$ \code{measure\_sigma\_c}}
\end{equation}
both fitted over the last two thirds of the run (dropping the initial adjustment,
during which the flow spins up from rest). $\sigma>0$ is an unstable meander;
$c>0$ means the bend train migrates downstream.

% ===================================================================
\section{What the model does \emph{not} contain, and what follows}
\label{sec:notcontain}

Strictly linear perturbation dynamics (no $\bm u'\!\cdot\!\nabla\bm u'$); no
rotation ($f=0$, a river); no vertical structure, hence no helical secondary flow
(depth-averaged away, i.e.\ this is the $A=0$ limit of \cite{ikeda1981}, whose
$A\approx2.89$ is their \emph{dominant} bend driver); the base state is frozen (its
steady residual is an implicit forcing); and the lateral viscosity uses the flat
rather than the exact curvilinear vector Laplacian --- a wall closure and short-wave
regulariser.

Three consequences must be stated plainly, because they bound what any result here
can mean.

\paragraph{(i) No sediment, therefore not the \emph{observed} wavelength selection.}
The bed $H$ is prescribed and only the banks move. Meander wavelength selection is
understood as a resonance between the forced bend flow and the \emph{free alternate-bar
mode of the sediment bed} \cite{blondeaux1985,seminara2006,camporeale2007}. With the bed
frozen that free mode does not exist. The driving is short-wave biased --- the
$C'=-\partial_s^2\zeta_c$ forcing scales as $k^2$ while the response scales as $k$ ---
so a preferred wavenumber can only come from whatever damps short waves.

It turns out something does, and it is not the artificial viscosity: measurements in
\S\ref{sec:results} give a $k^2$ damping coefficient about twenty-six times $\nu$, set by
the cross-channel shear, with $k_{\rm peak}\Delta\simeq1.10$. So the model does select a
scale --- by shear, not by bar--bend resonance. It is the wrong mechanism for a river,
and it disappears as $\Delta\to0$ (whereupon the peak retreats to the genuine viscous
cutoff), so no comparison with the observed $\lambda\approx10W$ \cite{leopold1960} is
warranted. The point of this subsection stands; only the reason does not.

\paragraph{(ii) The erodibility is not in the geomorphic regime.} Field calibration
gives $E/U\sim10^{-8}$ \cite{hasegawa1977,nanson1986}; the default here is
$O(10^{-1})$. Classical theories take $E\to0$, which is what makes the flow
quasi-steady and justifies imposing $\un'(\pm b)=0$ at a \emph{fixed} wall. At
$O(1)$ erodibility the neglected wall-normal velocity $E\us'(\pm b)$ is not small
compared with the retained $\un'$, so the boundary condition is inconsistent at
leading order, and $\sigma$, $c$ are properties of a strongly coupled flow--bank
oscillator rather than of a river.

\paragraph{(iii) Froude insensitivity is weak evidence by construction.} Writing
$P\equiv g\eta$, the momentum equations contain no $F$ at all and $F$ survives only
as $F^{2}$ multiplying the time-and-advection derivative of $P$ in continuity ---
i.e.\ purely as a divergence (compressibility) parameter. Any slow, balanced mode is
therefore $F$-insensitive as a matter of scaling, and the observed insensitivity
\emph{confirms} the rigid-lid reduction rather than discriminating between wave
families. Distinguishing a free vortical wave from a forced boundary mode requires
the diagnostics of \S\ref{sec:classify}, not a Froude sweep.

% ===================================================================
\section{Classifying the mode (what the diagnostics mean)}
\label{sec:classify}

\code{classify\_mode()} reports, from the perturbation fields:
\begin{align*}
  \|\delta'\|/\|\zeta'\| &\quad\text{divergent vs rotational: }\to0\text{ means balanced (not a gravity wave)},\\
  \|q'\|/\|\zeta'\| &\quad q'=\zeta'/\bar h-\bar\zeta\eta'/\bar h^{2}\text{: how much vortex stretching},\\
  \|\eta'\|/\|\bm u'\| &\quad\text{should scale as }F^{2}\text{ for a balanced mode},\\
  T_{\rm shear}&=-\!\int \us'\un'\,\dn\Us
   \quad\text{the ONLY channel by which the mean-flow PV gradient can power a wave},\\
  T_{\rm bend}&=-\!\int \un'\,\Us^{2}\,\partial_s^{2}\zeta_c
   \quad\text{work done on the fluid by the moving bank.}
\end{align*}
The decisive one is $T_{\rm shear}$. A free vortical (Rossby-type) wave is
\emph{defined} by extracting wave activity from the background PV gradient, so it
requires $T_{\rm shear}>0$. If instead $T_{\rm shear}\le0$ the mean flow is a
\emph{sink} and the disturbance is being driven by the boundary, whatever its Froude
behaviour. Note also that a smooth, single-signed $\dn\bar q$ (which is what a
parabolic jet gives) has no PV edge and therefore supports \emph{no discrete}
vortical eigenmode at all --- only a continuous spectrum \cite{case1960}; discrete
counter-propagating Rossby waves require localised PV gradients \cite{bretherton1966}.
Planetary $\beta$ supports Rossby waves with no mean flow whatever; a shear-$\beta$
does not, because it \emph{is} the mean flow.

% ===================================================================
\section{What the parameter study actually returns}
\label{sec:results}

\code{experiments.py} runs one broadband simulation per physical configuration,
varying the bed, the initial bank sinuosity, the friction and the two base speeds one
at a time. Table~\ref{tab:results} is generated directly from the output files by
\code{make\_results\_table.py} --- it is never typed by hand, so it cannot drift from
the data it describes.

\begin{table}[h]\centering
\input{results_table}
\caption{Fastest \emph{converged} mode ($\ge3$ e-foldings) per configuration, and the
energetics ratio. $T_{\rm shear}>0$ would mean the mean-flow vorticity gradient powers
the disturbance.}
\label{tab:results}
\end{table}

Four readings follow. The first kills the headline one might have hoped for; the second
and third are corrections to earlier versions of this section that the controls forced,
and are recorded as such rather than quietly rewritten.

\paragraph{The mean flow is a sink, not the power source.} Across all eleven runs the
correlation is exact: \textbf{every growing run has $T_{\rm shear}\le0$, and the single
run with $T_{\rm shear}>0$ is the one that decays} (the matched reversed-shear control,
$T_{\rm shear}/|T_{\rm bend}|=+0.42$, $\sigma(k{=}1.8)=-0.053$). Whenever this
disturbance grows it is \emph{losing} energy to the background vorticity gradient rather
than extracting it, so by the criterion of \S\ref{sec:classify} it is not a free vortical
(Rossby-type) wave; the energy comes from $T_{\rm bend}$, the work done on the fluid by
the moving bank. Only the \emph{sign} of this ratio should be quoted: its magnitude is
resolution-sensitive (the reference run gives $-3.16$ at $N_s=64$ and $-0.36$ at
$N_s=128$), because the diagnostic is evaluated on the final field and the resolved mode
mix changes with $N_s$.

\paragraph{The channel-$\beta$ is not necessary --- but it is not irrelevant either,
and an earlier version of this section got that wrong.} The profile
$\Us=U_0+\Delta(1-n^2/b^2)$ has only two free constants, so removing $\Delta$ must
change either the bank speed $U_0$ or the centre speed: \emph{no perfectly clean control
exists}. The first attempt varied $\Delta$ while also raising $U_0$ from $0.4$ to $1.0$,
and since the erosion law reads $\us$ \emph{at the bank}, that conflated ``removed the
channel-$\beta$'' with ``sped up the bank''. The confound inverts the answer. Holding
the bank speed fixed at the reference value and varying only $\Delta$, compared at the
fixed well-resolved wavenumber $k=1.8$:
\begin{center}
\begin{tabular}{lrl}
\hline
$\Delta$ & $\sigma(k{=}1.8)$ & \\
\hline
$+0.60$ (reference) & $+0.1649$ & grows\\
$+0.00$ ($\beta$ removed) & $+0.0599$ & grows, but $2.75\times$ \emph{slower}\\
$-0.30$ (reversed) & $-0.0526$ & \textbf{stable} (no mode converged)\\
\hline
\end{tabular}
\end{center}
What survives is the qualitative statement: with the background vorticity gradient
\emph{identically zero} the meander still grows ($\sigma>0$), so the mechanism does not
require $\beta$ and cannot be a $\beta$-restored wave. What does \emph{not} survive is
the claim that it grows \emph{faster} without $\beta$ --- that was the $U_0$ confound.
Removing $\beta$ weakens the instability by a factor $\simeq2.75$, and reversing the
shear at matched bank speed suppresses it altogether.

\paragraph{Wavelength selection: there is one, and it comes from the shear.} This too
corrects an earlier reading. Fitting $\sigma\simeq ak-bk^{2}$ gives $k_{\rm peak}=a/2b$,
and the measured damping coefficient is $b\approx0.079$ while $\nu=0.003$ --- the $k^2$
damping that produces the peak is some \textbf{twenty-six times larger than the explicit
viscosity}, so the cutoff is not viscous. It tracks the shear instead:
$b=0.079,\,0.038,\,0.033,\,-0.003$ for $\Delta=0.6,\,0.2,\,0.0,\,-0.6$ (reversed shear
\emph{anti}-damps), and the peak obeys
\begin{equation}
  \boxed{\;k_{\rm peak}\,\Delta\simeq1.10\;}
\end{equation}
to within $\pm3\%$ across a tenfold change in $C_f$, a flat-to-parabolic bed, and bank
sinuosity $0\to0.15$. The earlier argument --- ``$k_{\max}$ barely moves while $C_f$
changes tenfold, therefore it is the $\nu$ cutoff'' --- was under-determined, because
all six of those runs held $\Delta=0.6$; ``$\Delta$-set'' fits the same evidence, and
varying $\Delta$ discriminates. As $\Delta\to0$ the shear damping vanishes and the peak
retreats to a residual viscous-plus-drag cutoff ($b_0\approx0.011$, $k_{\rm peak}=4.30$).
So the model does select a length scale, but by cross-channel shear, not by the
bar--bend resonance of \S\ref{sec:notcontain}(i); a comparison with the observed
$\lambda\approx10W$ remains unwarranted.

\paragraph{Why the movies look untidy, and why that is not numerical noise.}
The perturbation field never settles into a single clean travelling wave, and it is
worth being precise about why, because the obvious reading --- ``the simulation is
noisy'' --- is wrong on both counts.

\emph{It is a wave packet, not noise.} Measured on the final field, $90\%$ of the
streamwise-velocity variance lives in about \textbf{seven neighbouring modes},
$k\simeq1.65$--$2.10$, an r.m.s.\ spectral width of only $\Delta k\approx0.15$. Seven
nearby wavenumbers with unequal phases beat against one another, which looks irregular
but is entirely deterministic; broadband noise would instead show a flat spectrum. The
packet width is even predictable: near its maximum $\sigma\simeq\sigma_{\max}-a(k-k_p)^2$
with $a\approx0.077$ measured from the table, so after a time $T$ only modes with
$a(k-k_p)^2T\lesssim1$ survive, i.e.\ $\Delta k\sim(aT)^{-1/2}\approx0.3$ at $T=120$ ---
consistent with what is measured. The packet therefore narrows only as $T^{-1/2}$, and
the $\code{max\_efold}$ cap stops the run long before it becomes monochromatic.

\emph{The grid is clean.} The energy in the top decade of resolved wavenumbers is
$\sim10^{-16}$ of the total: there is no aliasing pile-up at all. And refining $N_s$
from $64$ to $128$ at fixed $\nu$ changes $\sigma(k)$ by $0.0$--$0.1\%$ across the whole
band, leaving the peak at $k=1.80$ --- the structure is converged, not grid-generated.

\emph{What $N_s=128$ does fix is legibility.} The dominant wavelength is
$\lambda=2\pi/1.86\approx3.4$, which at $N_s=64$ is only $2.6$ grid points per
wavelength --- barely above the two-point Nyquist limit, so \emph{any} wave, however
clean, would render as one-pixel stripes. At $N_s=128$ it is $5.2$ points per
wavelength and draws as the wave it is. The production runs use $N_s=128$ for that
reason alone; the physics is unchanged.

\paragraph{A note on run length, because it sets the numbers.} Since all modes are
seeded together, two of them separated by $\Delta\sigma$ reach an amplitude ratio
$e^{\Delta\sigma T}$; once that exceeds $1/\varepsilon\approx4.5\times10^{15}$ (some
$36$ e-foldings) the slower one is buried in the faster one's FFT round-off and simply
reads back the \emph{leader's} growth rate. The e-folding gate of \S\ref{sec:seed}
cannot catch this, because round-off inherits the leader's growth and therefore looks
perfectly converged; it is visible instead as a spurious sawtooth in $\sigma(k)$, in
which $\sigma$ depends only on $m$ modulo a fixed integer. At a common $t_{\rm end}=120$
the slowest configuration here grew a harmless $13$ e-foldings while the plug-flow run
grew $58$, and a third of the latter's spectrum was noise. Runs are therefore stopped
when the \emph{leading} mode has grown \code{max\_efold}$=25$ e-foldings, or at
$t_{\rm end}$, whichever comes first ($\to$ \code{run\_ivp\_SW}): the six slow runs are
unaffected and the three fast ones stop at $t=36$--$80$. This ties the usable dynamic
range to the diagnostic rather than to how fast a configuration happens to grow, and it
is why the fast runs are integrated over a shorter interval than the slow ones.

\paragraph{What the model \emph{is}, then.} A forced morphodynamic boundary mode:
balanced --- the divergence ratio is $\|\delta'\|/\|\zeta'\|\simeq0.010$--$0.051$, and
$F$ barely matters, both consistent with the rigid-lid reduction --- but driven by the
erodible boundary rather than by the interior PV gradient. The growth rate vanishes with
the erodibility, $\sigma\propto E$: it is not an eigenvalue of the flow alone. The
cross-channel shear then shapes it, setting both the amplitude (a factor $2.75$ between
$\Delta=0.6$ and $\Delta=0$, and outright stability for $\Delta<0$ at matched bank speed)
and the selected wavenumber ($k_{\rm peak}\Delta\simeq1.10$) --- but it never
\emph{powers} it, since $T_{\rm shear}\le0$ in every growing run. The useful corollary is
a validation result: since the free surface is passive here, the rigid-lid packages in
\code{../} are justified for $F\lesssim0.9$.

\paragraph{A methodological note worth keeping.} Two of the four readings above are
corrections, and both were caught the same way --- not by re-deriving, but by refining a
numerical parameter and by adding a control that held fixed the variable which had been
riding along. The $U_0$/$\Delta$ confound was invisible at $N_s=64$ because the runs it
corrupted were exactly the runs whose $\sigma(k)$ had not converged; doubling $N_s$ made
their growth rates move ($\sigma_{\max}$: $0.74\to1.49$), which is what prompted the
control in the first place. A parameter that cannot be varied independently --- here,
because $\Us$ has only two constants --- should be treated as confounded by default and
reported with an explicitly matched comparison, not with each run's own optimum.

\begin{thebibliography}{99}
\bibitem{ikeda1981} Ikeda, S., Parker, G. \& Sawai, K. (1981) Bend theory of river
  meanders. Part 1. Linear development. \emph{J. Fluid Mech.} \textbf{112}, 363--377.
\bibitem{parker1982} Parker, G., Sawai, K. \& Ikeda, S. (1982) Bend theory of river
  meanders. Part 2. \emph{J. Fluid Mech.} \textbf{115}, 303--314.
\bibitem{blondeaux1985} Blondeaux, P. \& Seminara, G. (1985) A unified bar--bend
  theory of river meanders. \emph{J. Fluid Mech.} \textbf{157}, 449--470.
\bibitem{seminara2006} Seminara, G. (2006) Meanders. \emph{J. Fluid Mech.}
  \textbf{554}, 271--297.
\bibitem{zolezzi2001} Zolezzi, G. \& Seminara, G. (2001) Downstream and upstream
  influence in river meandering. \emph{J. Fluid Mech.} \textbf{438}, 183--211.
\bibitem{camporeale2007} Camporeale, C., Perona, P., Porporato, A. \& Ridolfi, L.
  (2007) Hierarchy of models for meandering rivers. \emph{Rev. Geophys.} \textbf{45}, RG1001.
\bibitem{smith1984} Smith, J. D. \& McLean, S. R. (1984) A model for flow in
  meandering streams. \emph{J. Geophys. Res.} \textbf{89}(C4), 7389--7408.
\bibitem{engelund1974} Engelund, F. (1974) Flow and bed topography in channel bends.
  \emph{J. Hydraul. Div. ASCE} \textbf{100}(HY11), 1631--1648.
\bibitem{johannesson1989} Johannesson, H. \& Parker, G. (1989) Linear theory of river
  meanders. \emph{River Meandering}, AGU Water Resour. Monogr. 12, 181--213.
\bibitem{colombini1987} Colombini, M., Seminara, G. \& Tubino, M. (1987)
  Finite-amplitude alternate bars. \emph{J. Fluid Mech.} \textbf{181}, 213--232.
\bibitem{hasegawa1977} Hasegawa, K. (1977) Computer simulation of the gradual
  migration of meandering channels. \emph{Proc. Hokkaido Br. JSCE}, 197--202.
\bibitem{nanson1986} Nanson, G. C. \& Hickin, E. J. (1986) A statistical examination
  of bank erosion and channel migration. \emph{Geol. Soc. Am. Bull.} \textbf{97}, 497--504.
\bibitem{parker2011} Parker, G. \emph{et al.} (2011) A new framework for modeling the
  migration of meandering rivers. \emph{Earth Surf. Process. Landforms} \textbf{36}, 70--86.
\bibitem{leopold1960} Leopold, L. B. \& Wolman, M. G. (1960) River meanders.
  \emph{Geol. Soc. Am. Bull.} \textbf{71}, 769--794.
\bibitem{case1960} Case, K. M. (1960) Stability of inviscid plane Couette flow.
  \emph{Phys. Fluids} \textbf{3}, 143--148.
\bibitem{bretherton1966} Bretherton, F. P. (1966) Baroclinic instability and the
  short wavelength cut-off in terms of potential vorticity. \emph{Q. J. R. Meteorol.
  Soc.} \textbf{92}, 335--345.
\end{thebibliography}

\end{document}
