% Variable-depth H(x,y): from absolute vorticity to potential vorticity.
% Rossby Palooza -- Alluvial Rivers & Jet streams.  Companion to dedalus_meander2/.
% Compile:  lualatex variable_h_pv.tex   (run twice for the footline)
\documentclass[aspectratio=169,10pt]{beamer}

\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{array}

\definecolor{cChannel}{HTML}{08519C}
\definecolor{cGrowth}{HTML}{238B45}
\definecolor{cVeloc}{HTML}{E6820A}
\definecolor{cCurv}{HTML}{6A51A3}
\definecolor{cEros}{HTML}{D7301F}
\definecolor{cBank}{HTML}{7F5539}
\definecolor{cMom}{HTML}{E6550D}

\usecolortheme{whale}
\setbeamercolor{structure}{fg=cChannel}
\setbeamercolor{frametitle}{fg=white,bg=cChannel}
\setbeamercolor{title}{fg=white,bg=cChannel}
\setbeamercolor{subtitle}{fg=cChannel}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{itemize items}[circle]
\setbeamerfont{frametitle}{size=\large}
\setbeamertemplate{footline}[frame number]

\newcommand{\hl}[2]{\textcolor{#1}{\textbf{#2}}}
\newcommand{\mhl}[2]{\textcolor{#1}{#2}}
\newcommand{\pp}{\psi'}
\newcommand{\Pp}{\Psi'}
\newcommand{\ub}{\bar u}
\newcommand{\zp}{\zeta'}
\newcommand{\zb}{\bar\zeta}
\newcommand{\Hb}{\bar H}
\newcommand{\bt}{\beta_{\mathrm{top}}}
\newcommand{\dt}{\partial_t}
\newcommand{\dx}{\partial_x}
\newcommand{\dy}{\partial_y}
\newcommand{\Fr}{\mathrm{Fr}}

\title{Variable Depth $H(x,y)$: Absolute Vorticity $\to$ Potential Vorticity}
\subtitle{dropping the flat-bed assumption -- the topographic-shear $\beta$}
\author{Rossby Palooza --- Alluvial Rivers \& Jet streams}
\date{Derivation companion to \texttt{numerical/dedalus\_meander2/}}

\begin{document}

\begin{frame}[plain]\titlepage\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{The one assumption we drop}
The constant-depth model (\texttt{dedalus\_meander}) assumed a \hl{cBank}{flat rigid
bed}, $H=\mathrm{const}$.  Dropping it (Station~1 assumption (iv)) changes the
\emph{conserved quantity}:
\[
\text{flat bed:}\quad \frac{D\zeta}{Dt}=\mathcal C \qquad\longrightarrow\qquad
\text{variable }H:\quad \hl{cChannel}{$\dfrac{D}{Dt}\!\left(\dfrac{\zeta}{H}\right)=\dfrac1H\,\mathrm{curl}\,\mathcal F$}
\]
Absolute-vorticity conservation becomes \hl{cChannel}{potential-vorticity (PV)
conservation}.  The Rossby restoring force gains a \hl{cCurv}{topographic-shear
$\beta$}.  Everything else -- jet, erodible banks, low-Froude rigid lid -- is kept.
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{Station 1 revisited: the mass-transport streamfunction}
Low-Froude rigid lid, but now $h\approx H(x,y)$ is \emph{not} constant.  Continuity
\[
\dt h + \nabla\!\cdot(h\mathbf U)=0 \;\xrightarrow{\ \Fr^2\ll1\ }\;
\hl{cChannel}{$\nabla\!\cdot(H\mathbf U)\approx 0$}
\]
So it is the \hl{cVeloc}{mass flux} $H\mathbf U$ that is non-divergent, not $\mathbf U$.
Introduce the \hl{cVeloc}{mass-transport streamfunction} $\Pp$:
\[
u=-\frac1H\,\dy\Pp,\qquad v=+\frac1H\,\dx\Pp,\qquad
\zp=\dx v-\dy u=\hl{cChannel}{$\nabla\!\cdot\!\Big(\tfrac1H\nabla\Pp\Big)$}
\]
The relative vorticity is now a \hl{cEros}{variable-coefficient elliptic} operator on
$\Pp$ (flat bed: $\zp=\nabla^2\Pp$).
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{The curl: the PV-tendency equation}
Curl of the shallow-water momentum equation (the surface term $-g\nabla\xi$ is a
gradient, killed exactly).  Using $\nabla\!\cdot\mathbf U=-\tfrac1H\mathbf U\!\cdot\!\nabla H$:
\[
\dt\zp+\mathbf U\!\cdot\!\nabla\zp-\frac{\zp}{H}\mathbf U\!\cdot\!\nabla H
=\mathrm{curl}\,\mathcal F
\quad\Longleftrightarrow\quad
\hl{cChannel}{$\dfrac{D}{Dt}\!\Big(\dfrac{\zp}{H}\Big)=\dfrac1H\,\mathrm{curl}\,\mathcal F$}
\]
The extra \hl{cCurv}{vortex-stretching term} $-\tfrac{\zp}{H}\mathbf U\!\cdot\!\nabla H$
is what the flat bed did not have.  It combines with base advection into a modified
$\beta$ -- next slide.
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{Linearize (base bed $\Hb(y)$): the SAME skeleton}
Base jet $\ub(y)$ over base depth $\Hb(y)$; $\zb=-\ub_y$.  The stretching term folds
into the advection, and the linear equation is \hl{cChannel}{term-for-term the flat-bed
model} with two replacements:
\[
\boxed{\;\dt\zp+\ub(y)\,\dx\zp+\hl{cCurv}{$\bt(y)$}\,\dx\Pp=\mathrm{curl}\,\mathcal F',\qquad
\zp=\nabla\!\cdot\!\Big(\tfrac1{\Hb}\nabla\Pp\Big)\;}
\]
\[
\hl{cCurv}{$\bt(y)=\dy\!\Big(\dfrac{\zb}{\Hb}\Big)=\dfrac{2D}{\Hb}+\dfrac{\ub_y\,\Hb_y}{\Hb^{2}}$}
\qquad(\text{flat bed }\Hb=1:\ \bt\to 2D,\ \zp\to\nabla^2\Pp)
\]
$\beta\to\bt$ (topographic-shear) and $\nabla^2\to$ the $\Hb$-weighted elliptic operator.
Flat bed \hl{cGrowth}{recovers \texttt{dedalus\_meander} exactly}.
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{An honest word on ``topographic $\beta$''}
In a \hl{cChannel}{non-rotating} river ($f=0$) the base PV is $\zb/\Hb$ -- relative
vorticity only.  Hence
\[
D=0\ \Rightarrow\ \zb=2Dy=0\ \Rightarrow\ \bt=0 \qquad
\hl{cEros}{(a bed bump alone makes NO $\beta$)}.
\]
The bed \emph{modulates the shear-$\beta$}; it does not create a standalone
topographic $\beta$ without planetary vorticity.  A formal knob $f_0$ (default $0$)
restores true \hl{cCurv}{topographic Rossby waves}:
\[
\bt=\dy\!\Big(\dfrac{\zb+f_0}{\Hb}\Big)=\underbrace{\dy(\zb/\Hb)}_{\text{shear}}
-\underbrace{f_0\,\Hb_y/\Hb^2}_{\text{topographic}}\qquad(f_0\neq0:\ \text{waves even at }D=0).
\]
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{Implementation: well-banded, and why}
A survey of the group's Dedalus code found \hl{cEros}{no variable-coefficient
elliptic-LHS precedent}; the one $\text{NCC}\times\text{unknown}$ on the LHS
destabilised the IMEX stepper, and the docs warn that rational $1/H$ NCCs make dense
matrices.  Safe, precedented route (as in the QG/shallow-water scripts):
\begin{itemize}
  \item carry $\zp$ as an \hl{cChannel}{auxiliary field} (first-order reduction);
  \item multiply the elliptic constraint through by $\Hb^{2}$ to clear the $1/H$:
  \[
  \hl{cGrowth}{$\Hb\,\nabla^2\Pp-\Hb_y\,\dy\Pp-\Hb^{2}\zp=0$}\qquad
  (\Hb\text{ polynomial}\Rightarrow\text{all-polynomial, well-banded});
  \]
  \item multiply the PV tendency through by $\Hb^2$ (eigenvalue-preserving row scaling).
\end{itemize}
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{The ``water outside the channel'': which 2nd-order effect?}
Drawing the linear field on the \emph{fixed} box $[-1,1]$ while the banks displace by
$\eta=\psi_b/U_0$ ($U_0<1\Rightarrow\eta>\pp$) leaves an $O(\text{ampl})$-wide strip
(``water not filling / water outside'').  It is \hl{cCurv}{purely geometric}, \emph{not} a
flow.  Four second-order things, routinely conflated:
\begin{itemize}\footnotesize
  \item \hl{cCurv}{Geometric $\delta_0=ke$} (intrinsic$\to$Cartesian): puts the field IN the
  meandering channel. \hl{cGrowth}{THIS is the gap}; \texttt{warp\_fill} is its leading order.
  Parker~1982 keeps $O(\delta_0^3)$.
  \item \hl{cMom}{Nonlinear $J(\pp,\zp)=O(\text{ampl}^2)$}: flow self-advection
  (fattening/skewing of the bend).  Dropped -- we are linear.
  \item \hl{cVeloc}{Ikeda $\nu=b/r_0$, kept to $O(\nu^2)$}: the FLOW's response to bend
  \emph{curvature} (superelevation $\propto F^2$, near-bank velocity), taken at
  $\delta_0\to0$.  A \emph{different parameter} -- curvature, not amplitude.
  \item \hl{cBank}{Secondary flow $A\approx2.89$}: 3-D helical $\partial u/\partial z$ that
  scours the outer bank.  Depth-averaged away in both models; Ikeda re-imports it as
  $\eta'/H=-A\,\mathscr C'\tilde n$.  We omit it (a bed/cross-section effect).
\end{itemize}
\hl{cEros}{The gap is filled by the coordinate map (warp), not by any flow.}  Ikeda's
$O(\nu^2)$ is curvature-flow at zero amplitude; secondary flow is 3-D bed-scour -- both are
\emph{different} things, orthogonal to this planform gap.
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{Along-channel bed: flow over a wavy bed}
A bed varying \emph{along} the channel, $H(x,y)=\Hb(y)[1+a_x\cos(k_b x)]$ (Ikeda's
alternate bars), breaks $x$-homogeneity -- Fourier modes couple, so it is
\hl{cEros}{IVP-only} (no normal modes).  The base flow is fixed by
\hl{cChannel}{discharge conservation} $\dx(Hu)=0$:
\[
Hu_0=\Hb(y)\ub(y)\ \Rightarrow\ \hl{cVeloc}{$u_0(x,y)=\dfrac{\Hb(y)\,\ub(y)}{H(x,y)}$}
\qquad(a_x{=}0:\ u_0=\ub,\ \text{the }\Hb(y)\text{ case}).
\]
Flow \hl{cVeloc}{speeds over bars, slows over pools}.  With prognostic PV $q'=\zp/H$ the
system stays polynomial-coefficient:
\[
H\dt q'+\Hb\ub\,\dx q'-\dy\Pp\,\dx\bar q_0+\dx\Pp\,\dy\bar q_0+\gamma q'=0,\quad
H\nabla^2\Pp-\dx H\,\dx\Pp-\dy H\,\dy\Pp-H^{3}q'=0.
\]
\hl{cCurv}{Bed$\leftrightarrow$bank resonance} (default $k_b=k$, bar wavelength $=$ meander
wavelength $=$ the Ikeda bar$\leftrightarrow$bend coupling): at $a_x{=}0.25$ the bed shifts
$\sigma\ 0.10\!\to\!0.11$, $c\ {-}0.26\!\to\!{-}0.19$ (enhanced growth, slower upstream march).
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{Verification ladder (all green)}
\begin{center}\small
\begin{tabular}{@{}lll@{}}
\toprule
check & what & result\\
\midrule
flat-bed reduction & variable-$H$ GEP vs \texttt{VL.channel\_modes} ($H{=}1$) & \hl{cGrowth}{$0.0$ (identical)}\\
$d3$ EVP vs GEP & flat + bumped bed, both closures & \hl{cGrowth}{$4.4\times10^{-9}$}\\
topographic Rossby & $f_0{=}0$: $\bt{=}0$; \ $f_0{\ne}0$: propagating wave & \hl{cGrowth}{as predicted}\\
IVP vs EVP & measured $\sigma,c$, flat + bumped bed & \hl{cGrowth}{$0.1\%$}\\
along-channel bed & \texttt{build\_ivp\_Hxy} at $a_x{=}0$ vs $\Hb(y)$ EVP & \hl{cGrowth}{$0.1\%$}\\
\bottomrule
\end{tabular}
\end{center}
\vskip4pt
Two independent discretisations (Dedalus Chebyshev spectral \& an FD+Richardson GEP)
agree to $\sim10^{-9}$ -- the $\Hb^2$-weighted PV reformulation is correct.
\end{frame}

% --------------------------------------------------------------------------- %
\begin{frame}{Result \& outlook}
\begin{itemize}
  \item A deeper \hl{cVeloc}{thalweg} $H(y)=1+a_H(1-y^2)$ shifts the growth band and the
  \hl{cMom}{group-velocity flip} $k_g$ (the upstream/downstream momentum boundary) to
  the right -- modestly, because a \emph{symmetric} bump partly cancels in $\bt$.
  \item \hl{cGrowth}{Along-channel bedforms $H(x,y)$ now done} (discharge-conserving base
  flow, previous slide): at bed$\leftrightarrow$bank resonance the bars measurably force the
  bends.  Bigger effects still expected from an \hl{cEros}{asymmetric} (point-bar) bed.
  \item Core is one Dedalus driver (\texttt{meander\_driver.py}); all figures/movies in
  \texttt{postprocessing/}.  Flat bed always recovers \texttt{dedalus\_meander}.
\end{itemize}
\end{frame}

\end{document}
